Approximation as Differentiation: Notes

### 2025-05-25 main/category-of-families.tex
Identify a bit of reorg needed for Grothendieck section.

The goal now is to equip the interpretation of a (first-order) program with a Galois connection relating
approximations of its inputs to approximations of its output. Since the specific lattices of approximations
depend on the value being approximated, the semantics of every type $\tau$ must have both a ``conventional''
component, providing the set $X$ which interprets $\tau$, plus an associated lattice of approximations
$\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$ with a single
free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have two
components: a function $f: X \to Y$ giving the conventional interpretation of $e$, plus a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

This pattern of objects and morphisms is captured by the construction used in CHAD~\cite{vákár22,nunes2023} to
describe automatic differentiation, namely the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$. For CHAD, $T(X)$ will be the functor category $\Func{X}{\FinVect}$, regarded
as the category of $X$-indexed families of finite vector spaces, with indexed linear maps; for Galois slicing
we will take $T(X) = \Func{X}{\LatGal}$, the category of $X$-indexed families of bounded lattices, with
indexed Galois connections as morphisms. The Grothendieck construction
$\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed category $T$, sometimes called the \emph{total
category} for $T$, is the category of all $\Set$-indexed families of objects of $T$, together with morphisms
that account for both structure within each family and how families transform under reindexing.

First we formalise $X$-indexed families and their categories.

@@

\begin{definition}
\label{def:Grothendieck}
Suppose an indexed category $T: \cat{C}^\op \to \Cat$. The \emph{Grothendieck construction} for $T$ is the
category $\Grothendieck{X}T(X)$ (also written $\Grothendieck{}T$) which has as objects, all pairs $(X,
\partial X)$ of an object $X$ of $\cat{C}$ and an object $\partial X$ of $T(X)$, and as morphisms $(X,
\partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of morphisms $f: X \to Y$ in $\cat{C}$ and
morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in $T(X)$.
\end{definition}

>>>>

The goal now is to equip the interpretation of a (first-order) program with a Galois connection relating
approximations of its inputs to approximations of its output. Since the specific lattices of approximations
depend on the value being approximated, the semantics of every type $\tau$ must have both a conventional
component as well as an approximating component, i.e.~a set $X$ which serves as the usual interpretation of
$\tau$, plus an associated lattice of approximations $\partial X(x)$ for every point $x \in X$. The semantics
of an expression $e$ of type $\sigma$ with a single free variable of type $\tau$ (with $\tau$ and $\sigma$
denoting $X$ and $Y$) must then also have two components: a function $f: X \to Y$ giving the conventional
interpretation of $e$, plus a family of Galois connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$
for every $x \in X$ which can be used for slicing over that expression.

This pattern of objects and morphisms is captured by the construction used in CHAD~\cite{vákár22,nunes2023} to
model automatic differentiation, namely the \emph{Grothendieck construction} for a particular set-indexed
category $T: \Set^\op \to \Cat$. For automatic differentiation, $T(X)$ will be the functor category
$\Func{X}{\FinVect}$, regarded as the category of $X$-indexed families of finite vector spaces, with indexed
linear maps as morphisms. For Galois slicing, we will take $T(X) = \Func{X}{\LatGal}$, the category of
$X$-indexed families of bounded lattices, with indexed Galois connections as morphisms.

The Grothendieck construction $\Grothendieck{}T$~(\defref{Grothendieck} below) for an indexed category $T$,
sometimes called the \emph{total category} for $T$, is the category of all set-indexed families of objects of
$T$, together with morphisms that account for both structure within each family and how families transform
under reindexing functions $f: X \to Y$. First we formalise $X$-indexed families and their categories.

@@

\begin{definition}
\label{def:Grothendieck}
\todo{Move earlier because doesn't assume $\cat{C} = \Set$.} Suppose an indexed category $T: \cat{C}^\op \to
\Cat$. The \emph{Grothendieck construction} for $T$ is the category $\Grothendieck{X}T(X)$ (also written
$\Grothendieck{}T$) which has as objects, all pairs $(X, \partial X)$ of an object $X$ of $\cat{C}$ and an
object $\partial X$ of $T(X)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial
f)$ of morphisms $f: X \to Y$ in $\cat{C}$ and morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in
$T(X)$.
\end{definition}
